Massively Parallel Pseudo-Spectral DNS and~LES for Particle-Laden Turbulent Flows under Two-Way Momentum Coupling

### 2023-01-31 msc-thesis.tex
completed self and azaba fixes to text; completes separation of figures and tables

Important comprehensive study of impact of subgrid-scale modelling on particle collision statistics in HIT was performed by \textcite{Jin2010}.
They used the pseudo-spectral method for the fluid flow calculations and their realisation of LES (i.e. using spectral eddy viscosity model) matches numerical method used in this thesis.
Their results confirmed adverse effect of LES on the accuracy of particle-pair statistics, as well as established relation between Stokes number, $St$, and sensitivity of these statistics to subgrid-scale modelling.
They concluded that such effects are most noticeable when $St < 3$.
This study involved only non-settling particles.
It was later complemented by \textcite{Rosa2017} that used the same numerical setup but focused on particles affected by gravity.
Their results confirmed previous conclusions regarding limitations of LES to properly represent collisional statistics, such as radial distribution function and radial relative velocity.
They also established influence of subgrid-scale modelling on seemingly single-particle statistics, such as average settling velocity.
All studies mentioned above were limited to one-way momentum coupling.

@@

In this study, that may be considered the continuation of \textcite{Rosa2017}, a similar setup for LES method is employed \parencite[based on earlier work by][]{Jin2010}.
In particular, the closure of the subgrid term is based on the concept of the spectral-eddy viscosity, $\nu_{e}$.
In spectral space, thefiltered velocity field $\hat{\bar{\mathbf{U}}}(\mathbf{k}, t)$ is governed by the following momentum equation \parencite{Pope2000}
\begin{equation}
\left\lbrace \frac{\partial}{\partial t} + [\nu + \nu_{e}(k|k_{c})]k^{2} \right\rbrace \hat{\bar{\mathbf{U}}}(\mathbf{k}, t) = \mathbf{P}(\mathbf{k}) \mathcal{F}(\bar{\mathbf{U}} \times \bar{\boldsymbol{\omega}}) + \hat{\mathbf{f}}(\mathbf{k}, t) ,
\label{eqn:les}
\end{equation}
where $\mathbf{P} \equiv \left\{ \mathbf{P}_{ij} \right\}$ is the projection tensor (see Equation \ref{eqn:proj-tensor}) and $\mathcal{F}$ -- Fourier transform.
By $k_{c}$ we denote cutoff wavenumber for filtering and its value is set using the same formula as maximal resolved wavenumber, $k_{\max}$, in DNS, i.e. $k_c = \left\lfloor (N - 3) / 2 \right\rfloor$ (for $N=64$, we have $k_c = 30$).
In this method of modelling turbulence a sharp spectral filter is used to separate filtered (resolved) and residual (modelled) velocity fields which are strictly separated from each other in the spectral space with $| \mathbf{k} | = k_c$ as the boundary.
Hence, modelled modes for $| \mathbf{k} | > k_c$ have no bearing on turbulence model for larger wavenumbers and all necessary information on filtered velocity field $\mathbf{\hat{\bar{U}}}(\mathbf{k}, t)$ is contained in wavenumbers smaller than $k_c$.
Conversely, filtered velocity field contains no information on residual (small-scale) motions, thus they have to be entirely modelled \parencite[p. 615]{Pope2000}.

In Equation \ref{eqn:les}, the term $\nu_{e}(k|k_{c})k^{2} \hat{\bar{\mathbf{U}}}(\mathbf{k}, t)$ represents the specific subgrid-scale model that is used for parameterisation of small-scale turbulence.
Here, following \textcite{Rosa2017}, a standard formula for the spectral eddy viscosity is used \parencite[see][]{Chollet1981}.
In particular, the eddy viscosity is given by
\begin{equation}
\nu_{e}(k|k_{c}) = \nu_{e}^{+}(k|k_{c}) \sqrt{\frac{E(k_{c})}{k_c}} ,
\label{eqn:sgs-1}
\end{equation}
where total spectral energy for threshold wavenumber, $E(k_c)$, is dynamically evaluated at each time step, while the remaining term depends on $k$ and is defined as follows
\begin{equation}
\nu_{e}^{+}(k|k_{c}) = C_{K}^{-3/2} [ 0.441 + 15.2 \exp( -3.03 k_{c} / k) ] ,
\label{eqn:sgs-2}
\end{equation}
where $C_{K}$ is the model constant (here, we set $C_{K} = 2.5$).

@@

NNN

>>>>

Important comprehensive study of impact of subgrid-scale modelling on particle collision statistics in HIT was performed by \textcite{Jin2010}.
They used the pseudo-spectral method for the fluid flow calculations and their realisation of LES (i.e. using spectral-eddy viscosity model) matches numerical method used in this thesis.
Their results confirmed adverse effect of LES on the accuracy of particle-pair statistics, as well as established relation between Stokes number, $St$, and sensitivity of these statistics to subgrid-scale modelling.
They concluded that such effects are most noticeable when $St < 3$.
This study involved only non-settling particles.
It was later complemented by \textcite{Rosa2017} that used the same numerical setup but focused on particles affected by gravity.
Their results confirmed previous conclusions regarding limitations of LES to properly represent collisional statistics, such as radial distribution function and radial relative velocity.
They also established influence of subgrid-scale modelling on seemingly single-particle statistics, such as average settling velocity.
All studies mentioned above were limited to one-way momentum coupling.

@@

In this study, that may be considered the continuation of \textcite{Rosa2017}, a similar setup for LES method is employed \parencite[based on earlier work by][]{Jin2010}.
In particular, the closure of the subgrid term is based on the concept of the spectral-eddy viscosity, $\nu_{e}$.
In spectral space, the filtered velocity field $\mathbf{\hat{\bar{U}}}(\mathbf{k}, t)$ is governed by the following momentum equation \parencite{Pope2000}
\begin{equation}
\left\lbrace \frac{\partial}{\partial t} + [\nu + \nu_{e}(k|k_{c})]k^{2} \right\rbrace \mathbf{\hat{\bar{U}}}(\mathbf{k}, t) = \mathbf{P}(\mathbf{k}) \mathcal{F}(\mathbf{\bar{U}} \times \boldsymbol{\bar{\omega}}) + \hat{\mathbf{f}}(\mathbf{k}, t) ,
\label{eqn:les}
\end{equation}
where $\mathbf{P} \equiv \left\{ \mathbf{P}_{ij} \right\}$ is the projection tensor (see Equation \ref{eqn:proj-tensor}) and $\mathcal{F}$ -- Fourier transform.
By $k_{c}$ we denote cutoff wavenumber for filtering and its value is set using the same formula as maximal resolved wavenumber, $k_{\max}$, in DNS, i.e. $k_c = \left\lfloor (N - 3) / 2 \right\rfloor$ (for $N=64$, we have $k_c = 30$).
In this method of modelling turbulence a sharp spectral filter is used to separate filtered (resolved) and residual (modelled) velocity fields which are strictly separated from each other in the spectral space with $| \mathbf{k} | = k_c$ as the boundary.
Hence, modelled modes for $| \mathbf{k} | > k_c$ have no bearing on turbulence model for larger wavenumbers and all necessary information on filtered velocity field $\mathbf{\hat{\bar{U}}}(\mathbf{k}, t)$ is contained in wavenumbers smaller than $k_c$.
Conversely, filtered velocity field contains no information on residual (small-scale) motions, thus they have to be entirely modelled \parencite[p. 615]{Pope2000}.

In Equation \ref{eqn:les}, the term $\nu_{e}(k|k_{c})k^{2} \mathbf{\hat{\bar{U}}}(\mathbf{k}, t)$ represents the specific subgrid-scale model that is used for parameterisation of small-scale turbulence.
Here, following \textcite{Rosa2017}, a standard formula for the spectral-eddy viscosity is used \parencite[see][]{Chollet1981}.
In particular, the eddy viscosity is given by
\begin{equation}
\nu_{e}(k|k_{c}) = \nu_{e}^{+}(k|k_{c}) \sqrt{\frac{E(k_{c})}{k_c}} ,
\label{eqn:sgs-1}
\end{equation}
where total spectral energy for threshold wavenumber, $E(k_c)$, is dynamically evaluated at each time step, while the remaining term depends on $k$ and is defined as follows
\begin{equation}
\nu_{e}^{+}(k|k_{c}) = C_{K}^{-3/2} [ 0.441 + 15.2 \exp( -3.03 k_{c} / k) ] ,
\label{eqn:sgs-2}
\end{equation}
where $C_{K}$ is the model constant (here, we set $C_{K} = 2.5$).

@@

In this Chapter, the issues of computational performance and comparison of DNS and LES on that field are addressed.
As introduced before, the main advantage of LES is its supposed edge in numerical efficiency due to the reduction of grid size on which the turbulent flow is directly resolved.
This is the benefit that may outweigh, at least in some contexts, inaccuracies reported in Chapter \ref{ch:ch2} that are inherent to LES due to its parameterisation of small-scale turbulence.

In the following sections a more detailed analysis of the solver code is presented, especially in the context of its massively parallel execution.
Also some previous results concerning its scaling with respect to the grid size and number of CPU cores used in execution shall be mentioned.
The core of this Chapter, however, is the presentation of experimental measurements of code performance (in terms of wall-clock time of its execution per time step) with respect to several parameters of interest.
In addition, more in-depth results---such as time spent on different stages of code, time used for communication between processes, and impact of the particle distribution on execution time---are analysed to validate and expand on previous observations.

\section{Structure of Massively Parallel Simulation Code. Notes on Scaling}
\label{sc:ch3.perf}

First, building up on information provided in Section \ref{sc:ch1.impl}, a more detailed view on the architecture of the pseudo-spectral solver shall be given here.
As stated before, it uses 2D decomposition to divide the entire computational domain into column-like sections, or subdomains (sometimes also referred to as ``pencils'') that are handled in parallel by individual processes with exclusively attributed blocks of memory.
This makes the issue of communication and data transfer between subdomains critical for performance.
The selection of most important execution parameters for simulations used to obtain data in Chapter~\ref{ch:ch2} and most of the data in this chapter (unless otherwise noted) is presented in Table \ref{tab:perfs-params}.

\begin{table}[h]
\centering
\scriptsize
\begin{tabular}{lrccrcrr}
 & $N$ & $N^3$ & Subdomain size & $\#\text{SD}$ & $N / \text{SD}$ & $N_{\text{proc}}$ & $\# \text{Nodes}$ \\ \hline
\textbf{DNS} & $256$ & $2^{24} \; (\sim 16.8 \, \text{M})$ & $16 \times 16 \times 256$ & $256$ & $2^{16}$ & $256$ & $11$ \\
\textbf{LES} & $64$ & $2^{18} \; (\sim 262 \, \text{k})$ & $8 \times 8 \times 64$ & $64$ & $2^{12}$ & $64$ & $3$ \\   
\end{tabular}
\caption{Standard execution parameters for DNS and LES code: $N$ -- number of grid nodes (in one spatial direction); $N^3$ -- total number of grid nodes; $\#\text{SD}$ -- number of subdomains; $N / \text{SD}$ -- number of grid nodes per subdomain; $N_{\text{proc}}$ -- number of parallel processes used ($= \#\text{SD}$); $\# \text{Nodes}$ -- number of \emph{Okeanos} supercomputer nodes (24 CPU cores per node) to fully saturate maximal supported number of processes.
}
\label{tab:perfs-params}
\end{table}

It is notable that LES provides immense reduction on all axes of parameters that influence the theoretical complexity of fluid flow simulation.
This method not only requires resolution of fluid velocity field for ${2^6 = 64}$ times less nodes in total, but also every subdomain (handled by separate process) has ${2^4 = 16}$ times less nodes to compute.
Furthermore, the overall number of subdomains (and, equivalently, separate processes used) is smaller for LES by the factor of $4$.
For supercomputer \emph{Okeanos} (ICM UW), where each computational node comprises of $24$ CPU cores, this leads to significant reduction of utilised computational nodes at one time from $11$ to $3$.

These parameters, however, are only related to the fluid flow computations and does not reflect the effort that is necessary to resolve particles and particle-fluid interactions in the system.
This is especially vital in simulations under two-way momentum coupling where particle-to-fluid ratio is of great importance, thus necessitating simulation with large numbers of individually tracked computational particles.
This issue has been already hinted at previously as this is the main reason for introducing the super-particle parameterisation to model systems that achieve considerably larger mass loadings.
Furthermore, it must be observed that code used for DNS and LES solvers is almost identical, barring additional LES code for computation of viscosity contribution from subgrid-scale model (see Stage 9 in Table \ref{tab:code-steps}).
This has some clear advantages, including better comparability of both methods and code maintainability.
On the other hand, the parallel implementation of DNS is focused on splitting fluid flow computations that become much less burdensome in LES and computations pertaining to particles tend to quickly dominate computational costs, as will be shown in the following sections.

To achieve better understanding of the solver code structure, Table \ref{tab:code-steps} provides the outline of a single time step being executed, divided into ten stages.
These stages are grouped into ones that are associated with fluid (marked by ``F'' and greenish colours; Stages 1 and 7--10) and particles (marked by ``P'' and brownish colours; Stages 2--6).
In addition, we distinguish stages that are not directly involved in simulation, but perform necessary postprocessing of fluid and particle statistics (marked by ``d''; Stages 5 and 7).
Note that fluid statistics are computed and stored only in certain intervals (here every $50$ steps when particles are involved and every $20$ steps for particle-free simulations; in general, the frequency of data dumps can be adapted to the specific case).
Finally, there are Stages 3 and 9, that are enabled only for TWC or LES simulations, respectively.

\begin{table}[!htbp]
\centering
\scriptsize
\begin{tabular}{p{10mm}p{75mm}p{25mm}p{27mm}}
\textbf{No.} \newline (Cat.) & \textbf{Name} \newline \emph{Description} & \textbf{Interdomain \newline communication} & \textbf{Notes} \\ \hline \hline
\rowcolor[RGB]{190,240,220} \textbf{1} \newline (F) & 
\textbf{Transform $\mathbf{\hat{u}}$ and $\boldsymbol{\hat{\omega}}$ to physical space} \newline \emph{Uses inverse 3D FFT twice; first two steps of pseudo-spectral method procedure.} & FFT & Eqn. \ref{eqn:psproc-1} and \ref{eqn:psproc-2}. \\ 
\rowcolor[RGB]{240,220,190} \textbf{2} \newline (P) & 
\textbf{Calc. momentum transfer from fluid to particles} \newline \emph{Finds closest grid points to particles and calculates momentum transfer using six-point Lagrange interpolation scheme (OWC).} & Local & --- \\ 
\rowcolor[RGB]{240,190,150} \textbf{3} \newline (P) &
\textbf{Calc. momentum transfer from particles to fluid} \newline \emph{Uses PNN kernel to apply momentum transfer projected from particles onto fluid at neighbouring grid points (TWC).} & Local & Eqn. \ref{eqn:twc-px}. \newline \textbf{(TWC only)} \\ 
\rowcolor[RGB]{240,220,190} \textbf{4} \newline (P) &
\textbf{Calc. new particle velocities and positions} \newline \emph{Integrates equations of motion for particles to obtain their new velocities and positions; updates data structures used to keep track of neighbouring particles.} & Local & Eqn. \ref{eqn:owc}. \\ 
\rowcolor[RGB]{240,240,170} \textbf{5} \newline (Pd) &
\textbf{Calculate and write particle collision statistics} \newline \emph{Detects and counts particle collisions ($\Gamma^D_{11}$) and relative particle-pair data within radial shells necessary to compute RDF and RRV.}  & Global (only to \newline write data), Local & Sec. \ref{ssc:ch2.coll.rdfrrv} and \ref{ssc:ch2.coll.gamma}; \newline every step;  for steps \newline $> 30 000$ (after \newline system relaxation). \\ 
\rowcolor[RGB]{240,220,190} \textbf{6} \newline (P) &
\textbf{Update particle velocities and positions} \newline \emph{Changes the state of particles as calculated in step 4; transfers data for particles crossing subdomain boundaries.} & Global, Local & --- \\ 
\rowcolor[RGB]{150,240,150} \textbf{7} \newline (Fd) &
\textbf{Calculate and write flow statistics and spectra} \newline \emph{Obtains and stores energy and dissipation spectra; calculates other base and derived flow statistics ($u'$ and $\epsilon$).} & Global, FFT, Local & Sec. \ref{ssc:ch2.flow.bstat} and \ref{ssc:ch2.flow.spec}; \newline every 50 steps*. \\ 
\rowcolor[RGB]{190,240,220} \textbf{8} \newline (F) & 
\textbf{Transform $\mathbf{u} \times \boldsymbol{\omega}$ to spectral space} \newline \emph{Uses 3D FFT to perform third step of pseudo-spectral method procedure; performs post-transform dealiasing.} & FFT & Eqn. \ref{eqn:psproc-3}. \\
\rowcolor[RGB]{150,190,240} \textbf{9} \newline (F) & 
\textbf{Calc. effective viscosity using subgrid-scale model} \newline \emph{Uses fluid energy spectrum and subgrid-scale model formulation to adjust kinematic viscosity with spectral-eddy viscosity, i.e. $\nu + \nu_e(k|k_c)$.} & Global & Eqn. \ref{eqn:sgs-1} and \ref{eqn:sgs-2}. \newline \textbf{(LES only)} \\ 
\rowcolor[RGB]{190,240,220} \textbf{10} \newline (F) &
\textbf{Integrate fluid flow equations} \newline \emph{Integrates 3D Navier-Stokes equation using Crank-Nicholson scheme (fourth step of pseudo-spectral method); incorporates deterministic forcing and other external forces; performs necessary dealiasing and symmetrisation.} & Global, Local & Eqn. \ref{eqn:psproc-4} \\ \hline \hline
\end{tabular}
\caption{
The outline of a single time step of the pseudo-spectral solver used in this study that was divided into ten stages for readability.
Labels in the first column indicate whether given stage is related to solving fluid flow (``F'') or particles (``P''), and whether it is concerned with gathering and postprocessing data (``d'').
The third column characterises the level of interdomain communication required at given stage (see text for details). \\
* -- for simulations without particles flow statistics are gathered more frequently (here, every 20 steps).
}
\label{tab:code-steps}
\end{table}

The second column of Table \ref{tab:code-steps} provides general name and overview of computations performed at given stage.
For more information on general procedure of the pseudo-spectral method, see Appendix \ref{app:psm}.
The outline of strategy used to calculate particle-pair statistics (Stage 5) was described in Sections \ref{ssc:ch2.coll.rdfrrv} and \ref{ssc:ch2.coll.gamma}.
More information on detailed implementation, involving linked lists and cell-index method may be found in \textcite{Allen1987} and \textcite{Onishi2013}.
As for LES- and TWC-specific stages, these computations are outlined in Sections \ref{sc:ch1.les} and \ref{sc:ch1.part}, respectively.
See ``Notes'' column for additional references.

\begin{figure}[h]
\centering
\includegraphics[width=9cm]{figures/3-01_halo.pdf}
\caption{
The halo region of the host process that overlaps with eight neighbouring subdomains in 2D decomposition.
The range of overlap is $10R = 20a$, which is the largest possible outer radius of spherical shells used to calculate kinematic statistics of particles for particle located at the subdomain boundary.
Arrows indicate the direction of data transfer.
Illustration based on Figure 3 in \textcite{Ayala2014}. 
}
\label{fig:halo}
\end{figure}

Finally, Table \ref{tab:code-steps} contains distilled information on the severity of communication between processes required in given stage.
Stages marked with ``Local'' involve only communication between eight neighbouring subdomains in the physical space.
This communication is for the most part due to necessary exchange of fluid velocity field data for grid nodes on the subdomain boundary or transfer of particle data when they pass through such boundaries (Stage 6).
This communication is performed using MPI functions: \texttt{MPI\_SEND}, \texttt{MPI\_RECV}, and \texttt{MPI\_SENDRECV}.
This is also an important part of the data exchange when postprocessing particle data in Stage 5, since counting of kinematic statistics involves detection of particle pairs in spherical shells of radii related to the particle size.
These shells may overlap subdomain boundaries, hence it is necessary to transfer states of particles from regions of surrounding subdomains that may be affected \parencite{Ayala2014}.
These extra regions are referred to as the \emph{halo regions}.
For illustration, see Figure \ref{fig:halo}.

The label ``Global'' refers to data exchange between most or all processes and requires much greater degree of synchronisation between processes.
It is associated with MPI functions: \texttt{MPI\_BCAST}, \texttt{MPI\_REDUCE}, and \texttt{MPI\_ALLREDUCE}.
It is often involved in stages that compute global simulation statistics.
In one case, the postprocessing of turbulence statistics (Stage 7) requires global data of the entire fluid velocity field.
On the other hand, the gathering of particle-pair statistics (Stage 5) is mostly local and requires only information on neighbouring particles in halo regions.
The only global aspect of its interprocess communication comes into play at the end when all data is transferred to the one process that is responsible for disk I/O operations.
Finally, the \texttt{MPI\_ALLREDUCE} calls in Stages 9 and 10 cover necessary summations over fluid energy spectrum in the Fourier space.
In former case, it is necessary to calculate and redistribute to all processes the spectral-eddy viscosity involved in LES-specific subgrid-scale modelling.
For the last stage, such global synchronisation is necessary due to application of large-scale forcing (implemented in a sequential manner).
Also, to a limited degree, the data reduction along single axis is necessary in the symmetrisation of fluid velocity after performing Fourier transforms \parencite{Ayala2014}. 
Thus, we may observe that such synchronisation in LES-specific code should not negatively impact entire execution flow, as another global communication is imposed soon after Stage 9 in DNS code as well.

The very particular category of communication involves performing 3D Fast Fourier Transforms over the entire domain  as part of the pseudo-spectral solver in Stages 1 and 8 (also, less frequently, when gathering flow statistics).
It is marked by ``FFT''.
As previously mentioned, this code utilises effective implementation of the 3D FFT for distributed domains introduced in \textcite{Dmitruk2001}, adjusted for 2D domain decomposition by \textcite{Ayala2013}.
It uses strategy known as the transpose method that exploits the fact that multidimensional FFT may be decomposed into sequence of nested 1D FFTs.
Three consecutive 1D FFTs are performed in the direction of subdomain columns, or ''pencils'', along which the domain is not subdivided to avoid interprocess communication during computation of the transform.
Of course, to obtain result equivalent to 3D FFT it is necessary to transpose entire velocity field between and after 1D FFT calls, so they are effectively computed along all three axes.
This approach has been shown to be scalable with number of CPU cores used \parencite{Dmitruk2001,Ayala2013}.
Still, this transposition incurs relatively high cost due to necessary data transfer involving all processes.

Note, that this study focuses on single grid size for DNS and LES.
Hence, the data necessary for a broader and more systematic analysis of code scalability with respect to $N$ or $N_{\text{proc}}$ (i.e. the number of CPU cores used) is not available.
Some analyses of code performance when number of CPU cores available is limited, i.e. when execution is starved for resources, is presented in the following section, but it cannot be taken as a replacement for proper scalability study.
Fortunately, this implementation of the pseudo-spectral method has already been thoroughly studied in this respect.
The study by \textcite{Ayala2014} stands out with in-depth analysis of scaling for simulations without TWC (but with modelling of hydrodynamic interactions between particles), as well as development of interesting tools and concepts for more analytical and hardware-independent investigation of scalability.

This thesis is more focused the impact of simulating particles on the overall performance.
This is crucial, since it appears to be the dominating factor for LES.
Moreover, the influence of TWC on computational complexity was not thoroughly surveyed yet.
\textcite[Fig. 1 therein]{Rosa2022} provides simple scalability study of TWC simulations with respect to the particle radius and system mass loading and it was intended to show the difficulty of effectively attaining high $\Phi_m$ in simulations with smaller particles.
Further analysis of these subjects is presented in the following sections.

\section{Results of Basic Performance Measurements}
\label{sc:ch3.perfs}

To analyse impact of particles on the performance of the pseudo-spectral code the time of execution (the so-called wall-clock time, WCT) of individual steps were measured.
These basic measurements were taken for a single process.
Since global data transfer between processes at the end of each time step (Stage 10) has synchronising effect on all processes the impact of such simplification is minimal.
For the remainder of this Chapter these timings, averaged over the entire run of simulation, are denoted by $T$.
All uncertainties listed and marked on plots in this and following sections are standard deviations of these measurements.
Such deviations are relatively large, but this is expected as measurements involve a certain degree of ``natural'' variability between time steps, e.g. longer processing times whenever flow statistics are computed (i.e. every $20$ or $50$ time steps).
When comparison between DNS and LES is of interest, results are also presented in abstract unit of workload, $(T \times N_{\text{proc}})$, obtained by multiplying average wall-clock time per step by the number of CPU cores used at the same time in parallel execution (expressed in the unit of ``CPU-seconds'').
This measurement may serve as an approximation of the average amount of work or utilised computational resources per time step.

\begin{figure}[h]
\centering
\includegraphics[width=13.5cm]{figures/3-02_pfsowc.pdf}
\caption{
Time-averaged wall-clock times per step by itself (left) and multiplied by number of parallel processes used to execute code (right, in log scale) for simulations under one-way momentum coupling (OWC) presented as a function of particle radii.
Both results for DNS and LES are presented, with (WG) and without (NG) gravity.
Horizontal dashed lines represent values for simulations without particles.
}
\label{fig:pfsowc}
\end{figure}

Figure \ref{fig:pfsowc} shows the dependence of execution time on particle radius for simulations under one-way momentum coupling and without particles.
It may be observed that the wall-clock time per time step increases monotonically with particle radius, both in DNS and LES, with and without gravity.
Moreover, in case of simulations with settling particles, the impact of larger droplet radii is much stronger.
In part, these results may be explained by the fact that size of the halo region used to compute two-point particle statistics becomes larger with particle radii.
This means that more data related to the state of neighbouring particles needs to be transferred to the host process.
This implementation detail may explain slight increase in computational cost, as in case of results without gravity, but hardly such an increase as observed when gravity is involved.
More importantly, the clear difference between performance of simulations with and without gravity, predominantly for high inertia droplets (${a > 40}$~$\upmu\text{m}$), cannot be explained in simple terms, since enabling gravity in the solver code should have no direct bearing on its computational complexity.
This effect shall be further examined in following sections.

Note, that some attention has already been given to the relation between droplet radii and simulation performance, but other aspects have been emphasised.
In \textcite[Fig. 8c therein]{Ayala2014} the theoretical scaling analysis of Hybrid DNS code with respect to the sizes of droplets was conducted.
The comparison was made between simulations where liquid water content, comparable to $\Phi_m$ in this study, was held constant.
Simulations using smaller droplets required more individual particles to be tracked, thus increasing execution time.
It should be mentioned that simulations in \parencite{Ayala2014} were carried out under one-way momentum coupling but with consideration of aerodynamic interactions between droplets. 
Similar results were shown in \textcite[Fig. 1b therein]{Rosa2022} with TWC.
More interestingly, in the same paper (see Fig. 1a therein) it was argued that when the same number of particles is used, the execution time is broadly independent of particle radii.
These results, however, were performed only for non-settling particles.

Furthermore, the comparison of results obtained for DNS and LES confirms expectations that were previously anticipated.
Firstly, when only particle-free fluid simulation is concerned, LES is much faster due to reduced grid size.
This is even more pronounced for $(T \times N_{\text{proc}})$ as LES was found to consume two orders of magnitude less computational resources.
On the other hand, results for LES exhibit much larger difference in execution times between flow-only simulations and those with particles.
For DNS the extra computational cost of simulating $2$ million particles (on average $\sim 7.8\text{k}$ per subdomain) is barely noticeable, while for LES processing of $800$~k particles ($12.5\text{k}$ per CPU core) is dominating the execution time.

\begin{figure}[h]
\centering
\includegraphics[width=13.5cm]{figures/3-03_pfstwc.pdf}
\caption{
Time-averaged wall-clock times per step by itself (left) and multiplied by number of parallel processes used to execute code (right, in log scale) for simulations under two-way momentum coupling (TWC, with gravity)as a function of  the number of computational particles. 
Results for both DNS and LES are presented.
}
\label{fig:pfstwc}
\end{figure}

The above observation is thoroughly confirmed by similar measurements made for simulations under two-way momentum coupling with varying number of particles (considering gravity).
These results are shown in Figure \ref{fig:pfstwc}.
When the number of particles is increased, execution times per step quickly become prohibitively large.
That is the reason behind very limited range (up to $5$ million) of particle counts in LES simulations and any larger mass loading may only be achieved via super-particle parameterisation.
This is in stark contrast with DNS where simulations with $20$ million computational particles could be easily performed.
For illustration, on the hardware used to run these simulations (\emph{Okeanos}) the total running time of LES with more than 5 million particles would easily exceed $2$ days (i.e. maximal time of a single job) while DNS with $20$ million particles (largest used) would be able to compute almost $2.5$ times more time steps in the same amount of wall-clock time.

This result, even though expected, is much more severe than anticipated.
Furthermore, given code and parallelisation architecture used in this study, LES is completely untenable for TWC simulations with even moderate mass loading.
One simple remedy for this issue is the use of super-particle model, as it is done here.
This, however, introduces the combination of two distinct parameterisations, i.e. super-particles and subgrid-scale model, whose interaction is not well known and renders veracity of such results fragile at best.
In the long run, the best solution would be to establish the parallelisation strategy for LES that relies not only on decomposition of grid used to compute fluid flow (as with current implementation tailored for DNS) but also for particles which clearly are the bottleneck.

On another note, results shown in Figure \ref{fig:pfstwc} corroborate claims of \textcite{Rosa2022} that scaling of TWC DNS with respect to the number of particles is close to linear.
Furthermore, the measurements for TWC (both DNS and LES) confirm that the computational effort increases with growing particle sizes.

It was already mentioned that since this study is confined to one specified grid size, $N$, proper scaling analysis with respect to $N$ and $N_\text{proc}$ is not feasible.
Instead, figure \ref{fig:pfspfn} shows wall-clock times for a range of DNS simulations in relation to maximal number of computational nodes on supercomputer \emph{Okeanos} (24 CPU cores per node) that were set as a hard limit for the running code to be able to utilise.
It is important to point out that these simulations still exhibit standard grid size and division into subdomains (see Table \ref{tab:perfs-params}).
Thus, optimal number of parallel processes executing the code remained unchanged.
These results, therefore, provide information on the performance of simulations when it is starved for resources, i.e. the ratio of processes executing in parallel and available CPU cores is more than $1$.

\begin{figure}[h]
\centering
\includegraphics[width=8cm]{figures/3-04_pfspfn.pdf}
\caption{
Time-averaged wall-clock times per step for DNS, depending on the limited number of computational nodes on \emph{Okeanos} supercomputer (24 CPU cores per node).
DNS supports up to $256$ parallel processes in execution, hence value of $11$ represents optimal saturation of processes with available CPU cores.
}
\label{fig:pfspfn}
\end{figure}

The number of nodes enough to saturate all processes for ``standard'' DNS is $11$ (see Table~\ref{tab:perfs-params}).
When the number of available nodes is limited, the execution time increases linearly within certain intervals that are separated by radical drops in the computational performance.
These radical declines in performance, somewhat expectedly, coincide with thresholds where the ratio of the number of processes, $N_{\text{proc}}$, to the number of available CPU cores, $N_{\text{CPU}} = 24 \times \#\text{Nodes}$, becomes larger than consequent whole numbers.
For example, for $6$ Nodes such ratio is approximately $1.78$, which indicates that, in simple terms, each CPU core serves one process, and than another, with some CPU cores to spare.
For $5$ Nodes, however, $N_{\text{proc}} / N_{\text{CPU}} \approx 2.13$, thus every CPU has to serve two processes, and then some still remain to be executed.
This code, as outlined in Table \ref{tab:code-steps}, requires significant degree of local and global synchronisation between processes, hence such remaining processes, even if few, put free CPU cores on hold waiting for all processes to complete. 
In addition, certain time overhead must be accounted for queuing software and operational systems of nodes to handle more frequent preemptions and context switches between processes.

\begin{table}[h]
\centering
\scriptsize
\begin{tabular}{lrrrr}
\textbf{LES} & \multicolumn{4}{c}{Wall-clock time per step [ms]} \\
& No particles & OWC, $N_{\text{part}} = 2 \text{M}$ & TWC, $N_{\text{part}} = 2 \text{M}$ & TWC, $N_{\text{part}} = 5 \text{M}$ \\ \hline
1 node utilised             & $112.90 \pm 22.71$ & $1250.07 \pm 81.70$ & $1281.39 \pm 80.65$ & $6407.73 \pm 303.50$ \\
2 node utilised             & $2.22 \pm 0.33$ & $462.62 \pm 30.51$ & $488.70 \pm 35.98$ & $3348.79 \pm 118.54$ \\
3 node utilised (saturated) & $1.77 \pm 0.34$ & $249.76 \pm 17.41$ & $260.68 \pm 19.34$ & $1914.37 \pm 90.85$ \\
\end{tabular}
\caption{Time-averaged wall-clock times per step for LES, depending on the limited number of computational nodes on \emph{Okeanos} supercomputer (24 CPU cores per node).
Consistent with Figure \ref{fig:pfspfn}.
}
\label{tab:perfs-pfnles}
\end{table}

For completeness, results of similar measurements for LES are shown in Table \ref{tab:perfs-pfnles}.
These results show similar pattern, although it is harder to discern due to limited range of nodes that can be utilised due to fewer subdomains.
The only outlier is the exorbitantly high value for simulation without particles when limited to only single node.
This, however, may be at least in part explained by aforementioned overhead due to process switching that seems to dominate otherwise very fast flow-only computations with LES.

In addition, both OWC and TWC simulations with the same number of particles ($2$ million) are considered.
These results clearly show that additional computations due to TWC (Stage 3) have barely noticeable influence on the execution times.
Thus, it is not TWC itself that makes simulations slower but the necessity of increasing the number of particles in the system to obtain physically significant results.

\begin{table}[h]
\centering
\scriptsize
\begin{tabular}{lrrrrrr}
\multicolumn{7}{c}{\textbf{Wall-clock time per step [ms]}} \\
 & & No particles & OWC & TWC & TWC & TWC \\ 
Subdomain size & $N / \text{SD}$ &  & $N_{\text{part}} = 2 \text{M}$ & $N_{\text{part}} = 2 \text{M}$ & $N_{\text{part}} = 5 \text{M}$ & $N_{\text{part}} = 20 \text{M}$ \\ \hline

DNS, $8 \times 8$ & $2^{14}$ & $24.6 \pm 7.4$ & $39.2 \pm 8.1$ & $41.9 \pm 7.7$ & $60.0 \pm 11.0$ & $2015.2 \pm 29.4$ \\
\textbf{DNS}, $\mathbf{16 \times 16}$ & $2^{16}$ & $58.3 \pm 10.2$ & $89.7 \pm 9.8$ & $97.6 \pm 9.4$ & $180.0 \pm 20.8$ & $872.5 \pm 76.2$ \\
DNS, $32 \times 32$ & $2^{18}$ & $192.0 \pm 43.2$ & $310.6 \pm 45.0$ & $317.8 \pm 44.0$ & $551.8 \pm 45.9$ & $3046.2 \pm 139.4$ \\ \hline
\textbf{LES}, $\mathbf{8 \times 8}$ & $2^{12}$ & $1.8 \pm 0.3$ & --- & $338.4 \pm 29.5$ & $2539.7 \pm 246.5$ & --- \\
LES, $16 \times 16$ & $2^{14}$ & $4.1 \pm 0.5$ & $1359.4 \pm 49.1$ & $1425.9 \pm 52.4$ & $9142.4 \pm 290.8$ & --- \\ \hline \hline
 & & & & & & \\
\multicolumn{7}{c}{$\mathbf{(}$\textbf{WCT [s]} $\mathbf{\times N_{\text{proc}})}$ \textbf{per step}} \\
 & & No particles & OWC & TWC & TWC & TWC \\ 
Subdomain size & $N_{\text{proc}}$ &  & $N_{\text{part}} = 2 \text{M}$ & $N_{\text{part}} = 2 \text{M}$ & $N_{\text{part}} = 5 \text{M}$ & $N_{\text{part}} = 20 \text{M}$ \\ \hline
